Initial estuarine response to the nutrient dense Piney Point wastewater discharge into Tampa Bay, Florida

### 2021-10-08 manu-draft.Rmd
scriveners corrections to lines 88 & 90

HRK Holdings, LLC (hereafter, HRK) acquired Piney Point in August 2006 through an administrative agreement with FDEP.  This agreement transferred responsibility of the site to HRK with the intention that any future uses must protect and be compatible with the integrity of stack closure and long-term care.  In 2011, HRK agreed to the storage of 1.5 million cubic yards of dredged material and seawater from Port Manatee to improve shipping capacity at the port (i.e., Berth 12 construction).  This material was added to an existing gypstack at Piney Point. Placement of the dredged material was suspected in compromising the liner integrity which led to an emergency discharge that released 169 million gallons of dredged saltwater slurry and 3.5 tons of nitrogen to receiving waters leading to Bishop Harbor. The dredging and deposit of slurry at Piney Point continued following structural fortifications to the holding stacks to ensure integrity with additional loadings.  HRK maintains ownership and responsibility of the site to present day with oversight by FDEP

@@

Established laboratory and field sample protocols for all survey methods were based on an [Interagency Monitoring Project Plan](https://drive.google.com/drive/u/0/folders/1oBGvjdve-Gpo4Kn3Ovn8a8-yVoP25eec) maintained by the TBEP in agreement with USEPA standards and those of the inter-agency partners.  To the extent possible, data quality objectives followed guidelines outlined in the TBEP Data Quality Management Plan [@tbep1620]. Many of the local partners also participate in the Southwest Florida [Regional Ambient Monitoring Program](https://tbep.org/our-work/boards-committees/technical-advisory-committee/#ramp) that ensures similar standards and protocols are followed in the collection of monitoring data, including routine cross-reference of samples between laboratories to check precision of measured values.  Discrete water quality samples were taken primarily from surface grabs by boat and processed by the respective laboratories of each participating agency.  For this paper, we focus on parameters related to the Tampa Bay nutrient management paradigm and the expected phytoplankton response from a dense, inorganic nitrogen plume entering the bay.  This included evaluation of total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), and chlorophyll-a (ug/L) concentrations.  Samples for pH, salinity (psu), temperature (^o C), and dissolved oxygen saturation (\%) were also evaluated given the role these parameters can have as indicators of wastewater contamination (pH), physical drivers of primary production (salinity, temperature), and indicators of primary production and respiration (dissolved oxygen, mg/L).  Overall, sample effort was variable given agency resources at the time of the discharge event and over the next few months.  As appropriate, water quality data were aggregated at the weekly scale and by major areas of interest (Figure \@ref(fig:map)a) given the hypothesized impacts adjacent to the discharge site.  These areas of interest were delineated based on proximity to Piney Point, the main shipping channel in the bay, inflow boundaries (e.g., little Manatee River), location of the Skyway Bridge at the mouth of Tampa Bay, and major segments of Tampa Bay used by TBEP for assessing programmatic goals.

>>>>

HRK Holdings, LLC (hereafter, HRK) acquired Piney Point in August 2006 through an administrative agreement with FDEP.  This agreement transferred responsibility of the site to HRK with the intention that any future uses must protect and be compatible with the integrity of stack closure and long-term care.  In 2011, HRK agreed to the storage of 1.5 million cubic yards of dredged material and seawater from Port Manatee to improve shipping capacity at the port (i.e., Berth 12 construction).  This material was added to an existing gypstack at Piney Point. Placement of the dredged material was suspected in compromising the liner integrity which led to an emergency discharge that released 169 million gallons of dredged saltwater slurry and 3.5 tons of nitrogen to receiving waters leading to Bishop Harbor. The dredging and deposit of slurry at Piney Point continued following structural fortifications to the holding stacks to ensure integrity with additional loadings.  HRK maintains ownership and responsibility of the site to present day with oversight by FDEP.

@@

Established laboratory and field sample protocols for all survey methods were based on an [Interagency Monitoring Project Plan](https://drive.google.com/drive/u/0/folders/1oBGvjdve-Gpo4Kn3Ovn8a8-yVoP25eec) maintained by the TBEP in agreement with USEPA standards and those of the inter-agency partners.  To the extent possible, data quality objectives followed guidelines outlined in the TBEP Data Quality Management Plan [@tbep1620]. Many of the local partners also participate in the Southwest Florida [Regional Ambient Monitoring Program](https://tbep.org/our-work/boards-committees/technical-advisory-committee/#ramp) that ensures similar standards and protocols are followed in the collection of monitoring data, including routine cross-reference of samples between laboratories to check precision of measured values.  Discrete water quality samples were taken primarily from surface grabs by boat and processed by the respective laboratories of each participating agency.  For this paper, we focus on parameters related to the Tampa Bay nutrient management paradigm and the expected phytoplankton response from a dense, inorganic nitrogen plume entering the bay.  This included evaluation of total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), and chlorophyll-a (ug/L) concentrations.  Samples for pH, salinity (psu), temperature (^o C), and dissolved oxygen saturation (\%) were also evaluated given the role these parameters can have as indicators of wastewater contamination (pH), physical drivers of primary production (salinity, temperature), and indicators of primary production and respiration (dissolved oxygen, mg/L).  Overall, sample effort was variable given agency resources at the time of the discharge event and over the next few months.  As appropriate, water quality data were aggregated at the weekly scale and by major areas of interest (Figure \@ref(fig:map)a) given the hypothesized impacts adjacent to the discharge site.  These areas of interest were delineated based on proximity to Piney Point, the main shipping channel in the bay, inflow boundaries (e.g., Little Manatee River), location of the Skyway Bridge at the mouth of Tampa Bay, and major segments of Tampa Bay used by TBEP for assessing programmatic goals.
